---
title: "The Effects of the NBA COVID Bubble on the NBA Playoffs"
subtitle: "A Case Study for Home-Court Advantage"
author: |
  | Michael Price and Jun Yan
  | michael.price@uconn.edu
  | Department of Statistics, University of Connecticut
date: "`r format(Sys.time(), '%d, %B, %Y')`"
documentclass: article
papersize: letter
fontsize: 11pt
bibliography: citations.bib
biblio-style: asa
output: bookdown::pdf_document2
header-includes:
   - \usepackage[margin=1in]{geometry}
toc: False
---
Abstract: 
The 2020 NBA playoffs were played inside of a bubble in Disney World because of
the COVID-19 pandemic. This meant that there would be no fans in attendance, 
games played on neutral courts and no traveling for teams, which in theory 
removes home-court advantage from the games. This setting has attracted a lot of
discussion as analysts and fans debated the possible effects it may have on the
outcome of games. Home-court advantage has historically played a huge role in 
NBA playoff series outcomes. The 2020 playoff provided a unique
opportunituy to study the effects of the bubble, and home-court
advantage by comparing the 2020 season with the seasons in the past
While many factors contribute to the outcome of games, points scored
is the deciding factor of who wins games so scoring is the primary focus 
of this study. The specific measures of interest are team scoring totals and 
team shooting percentage on two pointers, three pointers, and free throws. 
Comparing these measures for home teams and away teams in 2020 vs. 2017-2019
shows that the 2020 playoffs favored away teams more than normal, particularly
with two point shooting and total scoring.

```{r setup, echo = FALSE, message =  FALSE, warning = FALSE}
library(BSDA)
library(ggplot2)
knitr::opts_chunk$set(fig.width = 6, fig.height = 4, dpi = 300,
out.width = "90%", fig.align = "center", dev = "pdf")
```


# Introduction {#sec:intro}
Home-court advantage is often discussed in sports circles as a contributing 
factor to the outcome of games. It is well-known that the home team typically
benefits from some competitive edge from playing at their home-court, resulting
in a better chance of winning. Thus, the NBA playing the 2020 playoffs in a 
bubble brought a great deal of concern for fans, teams, journalists, and others. 
Steve Aschburner discusses details about anticipated effects in his article 
“How Orlando's Neutral Site Will Impact Teams, Referees and Games” @Aschburner. 
Aschburner shares concerns from former players, coaches and other experts about
the potential effects of removing home-court advantage. Aschburner notes that the
NBA did make attempts to recreate the effects by putting the "home” team logo on
the court and allowing the "home" team to play crowd noise and music, but most 
people doubted these small attempts would recreate a true playoff atmosphere. 
During the 2020 NBA playoffs, home teams only won about 48.2% of the games. This
is lower than normal, which Aschburner claims usually floats around 60%. This 
shift in the home team winning percentage surely indicates the opportunity for
thorough investigation.

So, what happened? Did the home teams fail to perform up to normal standards 
without the help of home-court advantage? Were away teams able to rise to 
the occasion and perform better not having to deal with the headache of going
on the road? This paper seeks to answer that using scoring totals and shooting
percentages as indicators of team performance. This will deepen understanding
of how home-court advantage affects home and away teams in the NBA.

This study is quite different from earlier NBA home-court advantage studies. 
By using the neutral site games of 2020 we will get to compare home and away
performance to a control. Typically, studies just compare home vs away 
performance. These studies don’t separate the effects of home-court advantage
into the specific effect on the home team and the specific effect on the away
team. They show that home teams outperform away teams, but not if this is a 
result of home teams overperforming or away teams underperforming because of 
home-court advantage. Some of these studies are reviewed in greater detail in
Section \@ref(sec:litrev) Literature Review. 

This study will compare home team performance in 2020 at a neutral site with 
no fans vs. 2017-19 playoffs with fans. Likewise, away team performance in 2020
at a neutral site with no fans vs. 2017-19 playoffs with fans. By comparing home
teams in 2020 to home teams in 2017-19 and away teams in 2020 to away teams in 
2017-19, I can add a new perspective to the field of research. This will allow for
a more accurate understanding of the effects of home-court advantage on home and
away teams in the NBA. We won’t just see that home-court advantage helps home 
teams outperform away teams, but this study will actually clearly separate the 
effects of home-court advantage on home teams and away teams performance individually.

9 statistical tests are run to understand the differences in 2020 vs. earlier 
years. First, whether or not the difference between home win percentage in 2020
and 2017-19 is statistically significant. To confirm the difference we see in the
winning percentage is truly significant. Then we can look for differences in home
scoring in 2020 vs 2017-2019. Similarly, we can do the same test, but for 
differences in away scoring in 2020 vs 2017-2019. Also, differences in team 
shooting(for two pointers, three pointers, and free throws) from 2020 vs 2017-2019
for both home and away teams. The results from these simple tests will bring a new
understanding of how home-court advantage is impacting games by altering the performance
of the home and away teams.

# Literature Review {#sec:litrev}
There is tons of published literature on the effects of home-court advantage.
Many NBA home-court advantage studies analyze the effects by studying shooting
percentages. @Kotecki examines the impact of home-court advantage on winning in the
NBA. Kotecki proves the existence of home-court advantage using performance-based
statistics, specifically field goal percentage, free throw percentage, and 
points. All of which he proves significantly improved from home-court advantage, 
indicating home-court helps teams play better. This conclusion comes from 
comparing home performance vs. away performance in games. @Cao studied the effects
of pressure on performance in the NBA. They use free throws as their measure of 
interest, and research home and away factors in the study. They expected that home
fans could distract and put pressure on opposing players to make free throws,
which would negatively impact away team shooting. Likewise, the home team
would see a positive effect on free throw shooting. However, they find insignificant
evidence that home status has a substantial impact on missing from the free throw 
line. @Harris use two point shots, three-point shots and free throws as measures
of interest to study home-court advantage. They find that two point shots are
the strongest predictor of home-court advantage. They suggest that evidence 
supports that home teams should try to shoot more two point shots and force
their opponent to take more two point shot attempts. This strategy will maximize
the benefits of home-court advantage and give them the best chance to win.

Some studies focus less on shooting and more on scoring differences and other metrics.
For example, @Greer focuses his study on the influence of spectator booing on 
home-court advantage in basketball. The three methods of performance used
in this study were scoring, violations and turnovers. This study was conducted 
using the men’s basketball programs at two large universities. The study finds that
social support, like booing, is an important contributor to home-court advantage. 
Greer explains, whether the influence is greater on visiting team performance or
referee calls is less clear. However, the data does seem to lean slightly in
favor of affecting visiting team performance. One study conducted by @Harville studied the
effect of home-court advantage using the 1991-1992 college basketball season. 
Unlike the NBA, it is not uncommon to have a few games played at neutral sites
during the college basketball season. This allowed them to construct two samples,
one of home teams and one of neutral teams. They formulate their study as a 
regression equation predicting the expected difference in score for home teams. 
They set up their study to find if the home teams won games by more points when 
they had home-court advantage vs. when playing at neutral court. This study 
succeeds in proving that home-court advantage exists.

There are also studies focused on the factors contributing to home-court advantage. 
@Carron1992 cite 4 main game location factors for home and away teams.
First, the crowd factor, which is the impact of fans cheering. Second, learning
factors, which they categorize as the advantage from home teams from playing at
a familiar venue. Third, travel factors, the idea that away teams may face 
fatigue and jet lag from traveling. Last, rule factors, which says that home teams
may benefit from some advantages in rules and officiating. They acknowledge that
these factors would all be removed if games were played at a neutral site even 
if one team was designated as “home team”. This study was reviewed a decade later
by @Carron2005. The 2005 review goes over the new findings from studies 
about the significance of these 4 game location factors. Since 1992 they found 
that results on these 4 factors are mixed. However, there is some evidence 
supporting crowd and travel factors impact games in the NBA. There is less 
evidence suggesting learning and rule factors impact the NBA. One interesting 
finding cited by @Carron2005 is that in the absence of crowds result 
in overall performance increases. 


```{r data, include=FALSE}
#Data Collection
game_log_2020 <- read.csv(file="../data/Game_Log_2020.csv", header=TRUE)
game_log_2017_2019 <- read.csv(file="../data/2017-2019_game_data.csv", header=TRUE)

#Data Cleaning
game_log_2020$W.L <- ifelse(game_log_2020$W.L=="W", 1, 0)
game_log_2017_2019$W.L <- ifelse(game_log_2017_2019$W.L=="W", 1, 0)

game_log_2020$Home.Away <- ifelse(game_log_2020$Home.Away == FALSE, "Home", "Away")
game_log_2017_2019$Home.Away <- ifelse(game_log_2017_2019$Home.Away == FALSE, "Home", "Away")

game_log_2020_home <- game_log_2020[game_log_2020$Home.Away=="Home",]
game_log_2020_away <- game_log_2020[game_log_2020$Home.Away=="Away",]
game_log_2017_2019_home <- game_log_2017_2019[game_log_2017_2019$Home.Away=="Home", ]
game_log_2017_2019_away <- game_log_2017_2019[game_log_2017_2019$Home.Away=="Away", ]

#combineddata
combined <- rbind(game_log_2020, game_log_2017_2019)
combined$Year <- as.factor(combined$Year)
combined <-  na.omit(combined)

years <- 2010:2020
winpct <- c(.671, .667, .679, .635, .562, .593, .674, .570, .707, .561, .482)
```

# Data {#sec:data}
Data was collected from the official NBA website. The main variables of interest
are whether or not the home team won, scoring totals for home and away teams, and
shooting percentages for home and away teams on two pointers, three pointers, and
free throws. These variables were very popular and frequently used in the related
literature discussed earlier. While many other measures could be used for measuring
the outcome of the game and team performance, scoring seemed to be the most 
important. The winner of a game is determined by who scores more points. So, 
naturally when studying the outcome of winning, changes in scoring and scoring 
efficiency are important to understand.

Home-court advantage is the basic idea that the home team is more likely to win.
So laying a foundation of typical home-court advantage is crucial. Before 
focusing on the 2017 to 2020 playoffs we can take a quick look at home team win
percentages since 2010. Notice in Figure \@ref(fig:Fig1), the 10 years before 2020, the home team
winning percentage ranged from around 0.56 to 0.7 and never dipped below .5. The 
2020 bubble broke this historic pattern dipping down below .5. Foreshadowing, 
the confirmation of the expectation that the effect of home-court advantage was removed
in the 2020 playoffs.

(ref:cap-Fig1) "Winning percentage of NBA home teams in the playoffs since 2010, the green line denotes .500"
```{r Fig1, echo=FALSE, fig.cap = " (ref:cap-Fig1)", fig.align='center'}
winyear <- cbind(winpct, years)
winyear <- as.data.frame(winyear)
highlight <- data.frame(winpct = .482, years = 2020)
ggplot(winyear, aes(x=years, y=winpct))+ 
      geom_point()+
      geom_point(data=highlight, 
             aes(x=years, y=winpct), 
             color='red')+
      scale_x_continuous(name = "Year", breaks = years, labels = years)+
      scale_y_continuous(limits = c(0.4, 0.8))+
      geom_hline(aes(yintercept = 0.5), color = "green")+
      geom_text(aes(label=winpct), hjust=0, vjust=-1, size=2)
```
   
Moving on to the main focus of the study, comparing 2020 to 2017-2019. First, 
a brief look at the distribution of scoring depicted in Figure \@ref(fig:Fig2).
The green histograms indicate home and red histograms indicate away scoring. The
top row is scoring for 2020 and the bottom is scoring for 2017-2019. All histograms
seem to be fairly normally distributed which is important for statistical analysis.
It’s hard to see any clear differences between years, but it’s important to look at
and understand the distribution of measures before moving onto the analysis.   

(ref:cap-Fig2) Shows the home(blue) and away (red) scoring distributions for 2020(bottom) and 2017-2019(top)"
```{r Fig2, echo=FALSE, fig.cap="(ref:cap-Fig2)", fig.align='center'}
ggplot(combined) + 
  geom_histogram(aes(x = PTS, y = ..density.., fill = Home.Away), bins = 10, binwidth = 5)+
  facet_grid(Year ~ Home.Away)+
  theme(legend.position = "none")
```
   
Figure \@ref(fig:Fig3) shows home shooting for two pointers, three pointers and
free throws for 2020(top) vs. 2017-19(bottom). The histograms appear to be fairly
similarly distributed between 2020 and 2017-19. Likewise, Figure \@ref(fig:Fig4),
shows the same data except for away teams. Just like Figure \@ref(fig:Fig3),
Figure \@ref(fig:Fig4) histograms don’t show any obvious differences. The normality
of the distributions in both Figure \@ref(fig:Fig3) and Figure \@ref(fig:Fig4) are
important for meeting the required assumptions for statistical analysis.

(ref:cap-Fig3) Home shooting percentages for two pointers, three pointers and free throws for 2020(top) vs. 2017-19 (bottom)
```{r Fig3, echo=FALSE, fig.cap="(ref:cap-Fig3)"}
hometwo2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Two Pointers", 83),
                          Percentage=game_log_2020_home$X2P.) 
homethree2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Three Pointers", 83),
                            Percentage=game_log_2020_home$X3P.) 
homeFT2020 <- data.frame(Year=game_log_2020_home$Year, Type=rep("Free Throws", 83),
                         Percentage=game_log_2020_home$FT.) 
hometwo2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Two Pointers", 243),
                               Percentage=game_log_2017_2019_home$X2P.) 
homethree2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Three Pointers", 243),
                                 Percentage=game_log_2017_2019_home$X3P.) 
homeFT2017_2019 <- data.frame(Year=game_log_2017_2019_home$Year, Type=rep("Free Throws", 243),
                              Percentage=game_log_2017_2019_home$FT.) 
homegraph<-rbind(hometwo2020, homethree2020, homeFT2020, hometwo2017_2019, homethree2017_2019,
                 homeFT2017_2019)
                 
ggplot(homegraph) + 
  geom_histogram(aes(x = Percentage, y = ..density..), fill = "#00BFC4", bins = 10, binwidth = 5)+
  facet_grid(Year ~ Type)
```
  
(ref:cap-Fig4) Away shooting percentages for two pointers, three pointers and free throws for 2020(top) vs. 2017-19 (bottom)"
```{r Fig4, echo=FALSE, fig.cap="(ref:cap-Fig4)", fig.align='center'}
awaytwo2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Two Pointers", 83),
                          Percentage=game_log_2020_away$X2P.) 
awaythree2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Three Pointers", 83),
                            Percentage=game_log_2020_away$X3P.) 
awayFT2020 <- data.frame(Year=game_log_2020_away$Year, Type=rep("Free Throws", 83),
                         Percentage=game_log_2020_away$FT.) 
awaytwo2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Two Pointers", 243),
                               Percentage=game_log_2017_2019_away$X2P.) 
awaythree2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Three Pointers", 243),
                                 Percentage=game_log_2017_2019_away$X3P.) 
awayFT2017_2019 <- data.frame(Year=game_log_2017_2019_away$Year, Type=rep("Free Throws", 243),
                              Percentage=game_log_2017_2019_away$FT.) 
awaygraph<-rbind(awaytwo2020, awaythree2020, awayFT2020, awaytwo2017_2019, awaythree2017_2019,
                 awayFT2017_2019)
                 
ggplot(awaygraph) + 
  geom_histogram(aes(x = Percentage, y = ..density..), fill = "#F8766D", bins = 10, binwidth = 5)+
  facet_grid(Year ~ Type)
```
  
  Given the approximate normality of data, 9 different z-tests were performed in 
  order to test the effects of the COVID bubble on the 2020 NBA playoffs. The 9 questions were:  
* Is the home team winning percentage in 2020 different than that it was in 2017-2019?  
* Is the average home team scoring different in 2020 than it was over 2017-2019?  
* Is the average away team scoring different in 2020 than it was over 2017-2019?  
* Are home teams making two pointers at the same rate in 2020 as 2017-2019?  
* Are home teams making three pointers at the same rate in 2020 as 2017-2019?  
* Are home teams making free throws at the same rate in 2020 as 2017-2019?  
* Are away teams making two pointers at the same rate in 2020 as 2017-2019?  
* Are away teams making three pointers at the same rate in 2020 as 2017-2019?  
* Are away teams making free throws at the same rate in 2020 as 2017-2019?  

# Methods {#sec:methods}
The 2020 bubble provides a new and exciting opportunity to study home-court
advantage for the NBA. Unlike college basketball, aside from a few 
exhibition/preseason games, the NBA always has a home and away team. So, for
the first time in NBA history the bubble allows NBA home and away performance
to be compared vs a control/neutral field. The NBA bubble, as a neutral court,
removed all 4 possible game location factors impacting home-court advantage 
hypothesized by @Carron1992. The NBA bubble featured 8 seeding games then a
standard playoff format. The focus of this study was the play during the playoff
games since it followed the standard playoff format and can easily be compared
back to other playoffs. For this study, the 2020 playoffs were compared against
the three previous playoffs collectively. It is important to make sure that any
observed differences are truly from a home-court advantage. So, in order to control
for the changing play style of the NBA, we will limit the study to 2020 vs 2017-2019.
The reason being the faster pace play and more common use of the three-point shot 
in modern basketball. If we used data from say 10 years ago, or earlier, observed
differences may not be from effects of the NBA bubble, but rather from the effects
of drastic changes in the style of play between the seasons. However, basketball
evolves slow enough that we can reasonably assume 2017-2019 are at least very 
close in pace and playing style to 2020. 

Comparisons between 2020 and 2017-19 home and away teams were made on home team
winning percentage, total team scoring and two-point, three-point and free throw
shooting. Comparing the differences in these metrics for home and away teams in
2020 vs previous years will provide valuable insights to the understanding of 
home-court advantage. We can see how going on the road may negatively impact 
away performance and how playing at home may positively impact home performance.
If there are differences in scoring for home or away teams the differences can
be used to show how home-court advantage affects overall performance of home
and away teams. While testing for differences in shooting will provide added
context for how home-court advantage specifically affects performance. 
Shooting percentages aren’t the only possible metrics affected by home-court
advantage, but they are the most obvious and certainly an important one. In
order to compare these differences simple two-sample z-tests were conducted.
Test for differences in means was conducted for comparing scoring totals, shown in 
Equation \@ref(eq:zmean). Tests for differences in proportions were conducted for
comparing winning percentage and the three different shooting percentages, shown
in Equation \@ref(eq:zprop). All 9 tests are conducted under the assumption and null
hypothesis that there should be no difference in statistics from 2020 compared
to those from 2017-2019.

\begin{align}
    z = \frac{(\bar{X}_1 - \bar{X}_2)-0}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}.
    (\#eq:zmean)
\end{align}

\begin{align}
    z = \frac{(\hat{p}_1 - \hat{p}_2)-0}{\sqrt{\hat{p}(1-\hat{p})(\frac{1}{n_1}+\frac{1}{n_2})}}.
    (\#eq:zprop)
\end{align}

# Analysis and Results {#sec:results}
```{r, echo=FALSE}
#Analysis
one <- prop.test(x=c(sum(game_log_2020_home$W.L), sum(game_log_2017_2019_home$W.L)), 
                 n=c(nrow(game_log_2020_home), nrow(game_log_2017_2019_home)), correct=FALSE)
rowone <- c(as.numeric(one$estimate[1]), as.numeric(one$estimate[2]), one$conf.int,
          sqrt(as.numeric(one$statistic)), one$p.value)

two <- z.test(x=game_log_2020_home$PTS, y=game_log_2017_2019_home$PTS, sigma.x = 
                sd(game_log_2020_home$PTS), sigma.y = sd(game_log_2017_2019_home$PTS))
rowtwo <- c(as.numeric(two$estimate[1])/100, as.numeric(two$estimate[2])/100, 
            two$conf.int, two$statistic, two$p.value)

three <- z.test(x=game_log_2020_away$PTS, y=game_log_2017_2019_away$PTS, sigma.x = 
                  sd(game_log_2020_away$PTS), sigma.y = sd(game_log_2017_2019_away$PTS))
rowthree <- c(as.numeric(three$estimate[1])/100, as.numeric(three$estimate[2])/100, 
              three$conf.int, three$statistic, three$p.value)

four <- prop.test(x=c(sum(game_log_2020_home$X2PM), sum(game_log_2017_2019_home$X2PM)), 
                  n=c(sum(game_log_2020_home$X2PA), sum(game_log_2017_2019_home$X2PA)))
rowfour <- c(as.numeric(four$estimate[1]), as.numeric(four$estimate[2]), 
             four$conf.int, sqrt(as.numeric(four$statistic)), four$p.value)

five <- prop.test(x=c(sum(game_log_2020_home$X3PM), sum(game_log_2017_2019_home$X3PM)), 
                  n=c(sum(game_log_2020_home$X3PA), sum(game_log_2017_2019_home$X3PA)))
rowfive <- c(as.numeric(five$estimate[1]), as.numeric(five$estimate[2]), 
             five$conf.int,  sqrt(as.numeric(five$statistic)), five$p.value)

six <- prop.test(x=c(sum(game_log_2020_home$FTM), sum(game_log_2017_2019_home$FTM)), 
                 n=c(sum(game_log_2020_home$FTA), sum(game_log_2017_2019_home$FTA)))
rowsix <- c(as.numeric(six$estimate[1]), as.numeric(six$estimate[2]), six$conf.int,
           sqrt(as.numeric(six$statistic)), six$p.value)

seven <- prop.test(x=c(sum(game_log_2020_away$X2PM), sum(game_log_2017_2019_away$X2PM)), 
                   n=c(sum(game_log_2020_away$X2PA), sum(game_log_2017_2019_away$X2PA)))
rowseven <- c(as.numeric(seven$estimate[1]), as.numeric(seven$estimate[2]), 
              seven$conf.int,  sqrt(as.numeric(seven$statistic)), seven$p.value)

eight <- prop.test(x=c(sum(game_log_2020_away$X3PM), sum(game_log_2017_2019_away$X3PM)), 
                   n=c(sum(game_log_2020_away$X3PA), sum(game_log_2017_2019_away$X3PA)))
roweight <- c(as.numeric(eight$estimate[1]), as.numeric(eight$estimate[2]), 
              eight$conf.int,  sqrt(as.numeric(eight$statistic)), eight$p.value)

nine <- prop.test(x=c(sum(game_log_2020_away$FTM), sum(game_log_2017_2019_away$FTM)), 
                  n=c(sum(game_log_2020_away$FTA), sum(game_log_2017_2019_away$FTA)))
rownine <- c(as.numeric(nine$estimate[1]), as.numeric(nine$estimate[2]), 
             nine$conf.int,  sqrt(as.numeric(nine$statistic)), nine$p.value)

titles <- c(2020, "2017-19", "95% CI LL", "95% CI UL", "Test Stat.", "P-value")
rows <- c("Home Win %", "Home Scoring", "Away Scoring", "Home 2P%", "Home 3P%",
          "Home FT%", "Away 2P%", "Away 3P%", "Away FT%")
results <- data.frame(rowone, rowtwo, rowthree, rowfour, rowfive, rowsix,
                    rowseven, roweight, rownine, stringsAsFactors = FALSE)
resultsfinal <- data.frame(t(results))
resultsfinal1 <- round(resultsfinal[1:5], 3)
resultsfinal2 <- round(resultsfinal[6], 4)
resultsfinal <-cbind(resultsfinal1, resultsfinal2)
colnames(resultsfinal) <- titles
row.names(resultsfinal) <- rows
```

```{r table, echo = FALSE}
knitr::kable(resultsfinal, caption = 'The results from the 9 tests', align = "cccccc", booktabs = TRUE)
```

Starting from the top of the Table \@ref(tab:table), we see a statistically significant change
in home win percentage in 2020 from 2017-19, with p-value of 0.0365. The 95% 
confidence interval of (-0.255, -0.008) confirms our belief that home-court 
advantage was lost in the 2020 NBA playoffs. However, after accounting for 
multiple tests using Bonferroni correction the p-value is no longer significant.
So, we may only cautiously say there is evidence that home-court advantage was 
not a factor in 2020.

Home team performance did not seem to be negatively impacted by losing home-court
advantage like expected. Home scoring, two point, three point shooting and free 
throw shooting all show no significant difference, on average, between 2020 vs. 
2017-19. Home teams do not appear to have any evidence suggesting they played at 
a lower level in 2020 than they did in previous years when they had home-court advantage.

Away teams saw more of an impact than home teams. For starters, they saw a 
significant increase in mean points per game by about 5 points (95% CI: 2.083, 7.988)
Likewise, the away team two point shooting efficiency increased by an average of
about 0.03 (95% CI: 0.015, 0.050). However, unlike two point shooting, away 
teams did not see a statistically significant difference in three point and free
throw shooting. Overall, away teams have evidence of change in performance in 
the bubble. The away teams seemed to perform better than they would under normal
conditions as a visiting team.

# Discussion {#sec:disc}
Generally it seemed that away teams fared better in the 2020 NBA playoff bubble
than previous years on the road. Starting from the dip in home winning percentage
to below .482 it’s clear something was different. Although the difference was not
significant after a Bonferroni correction it’s still important to consider and
understand that home teams seemed to struggle to win compared to normal conditions.
Compared to @Kotecki who finds home teams consistently have a significantly
better record than away teams boasting about a 60.5% win percentage in his sample,
the 48.2% home winning percentage of 2020 home teams is a shift from normal.
It’s clear that home teams did not benefit from the usual advantages provided by
being the home team.

While the home teams in 2020 did see a slight uptick in average scoring it was
not large enough to be statistically significant. However, away team average 
scoring did increase by a statistically significant amount. This goes hand in 
hand with our intuition and conclusion about the home winning percentage 
decreasing. If away teams are scoring significantly more and home teams are not,
then we expect to see away teams winning a larger amount of games. This may give
more reason to believe the conclusion that there was a significant decrease in
home winning percentage in 2020, despite failing to be significant after Bonferroni
correction. Only away team scoring being significantly impacted by playing on
a neutral court indicates that home-court advantage stems mainly from adverse
effects on the visiting team.

An interesting finding is all shooting and scoring numbers for both home and
away teams did make at least small increases. Although these increases weren’t
all significant these increases are exactly what is reported in @Carron2005
when they explain how evidence suggests that teams perform better with the absence
of fans. This is important because it coincides with our conclusion that home-court
advantage mostly plays into games by negatively impacting away teams. If fans cause
overall performance to drop, then home court advantage must come from a bigger drop
in away performance than the drop in home performance. This is why away teams were 
able to close the gap with home teams with home-court advantage removed. 

Separating the effects of the home-court advantage into home effects and away
effects allowed for some interesting new insights. Previously, we knew that on
average home teams outperformed away teams. It was less clear whether it was 
from positive effects on the home team or negative effects on the road team or
perhaps a bit of both. The biggest takeaway from this study is the main source
of home-court-advantage is the negative effects playing on the road away teams 
face. In 2020 there wasn’t any evidence of regression for home team performance,
based on the performance measures used, despite being stripped of home-court
advantage. Yet, home teams lost about 12% more of games in the 2020 playoffs 
than the typical average. This was because of the improvement of away teams. 
No longer having to face the struggle of traveling, pressure from opposing fans,
or playing on an unfamiliar court, teams saw an improvement in their play and an
increase in winning. The improvement of away teams confirms a proposition from 
@Greer who believed that the positive social impact of crowds benefiting 
home teams may be a result of inhibiting away teams. 

At least some of that improvement from away teams came from significantly higher 
two point efficiency. This corresponds with the conclusion from @Harris,
where they found home teams are best suited to capitalize on advantages from two point
shots. Normally, by shooting more two pointers themselves and forcing away teams to 
shoot more two pointers the home team benefits from increasing effects of
home-court advantage. However, with away teams significantly improving two point
shooting in the bubble this strategy was no longer viable and home-court
advantage disappeared.

Future studies may want to use the 2020 NBA bubble and compare vs previous years
using other performance measures. For example, turnovers, steals, assist, rebounds,
and many more game statistics. There are plenty of other possibilities besides
just shooting efficiency to pick through looking for more possible sources of
added points for away teams. This will further help explain what is lost in the
performance of away teams when they travel to opposing arenas. This study is only
the beginning of possibilities for studies using the 2020 NBA bubble as a case study 
for home-court advantage.

Unfortunately, this study is limited by a one time sample. It seems unlikely that 
these conditions will ever be repeated. That being said, it will not be possible
to have a follow-up study using the same measures with a different sample. If
possible, that type of study could help strengthen the conclusions
made in this paper.

# References {#sec:ref}
